---
title: "SMC for random preferences"
output: rmarkdown::html_vignette
bibliography: bibliography.bib
vignette: >
  %\VignetteIndexEntry{SMC for random preferences}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(RanCh)
library(kableExtra)
library(corrplot)
library(tidyverse)
```

## Introduction

This vignette illustrates the use of tools for Sequential Monte Carlo (SMC) analysis of Bayesian random choice (BRC) and Bayesian random preference (BRP) models.
A detailed theoretical discussion of the SMC tools used here is in @McCaMarl13.
We illustrate here a small, but not trivial problem, so that the computations in this vignette take only a few seconds.
WJM: point to github stuff on full experiment.

## Preparing for simulation

All of the illustrations use data from @Tver72a, a paper that introduces both the elimination by aspects model and the "similary hypothesis", now usually known as the similary effect.
The universe of objects is the set of three choice objects of choice task $C$ (Gambles), denoted $T = \{x,y,z\}$.
Object $x$ is the lottery where one wins 4 Israeli pounds with probability $\tfrac{1}{5}$ (and nothing otherwise);
$y$ is the lottery where one wins 3.5 pounds with probability $\tfrac{1}{4}$; and $z$ is the lottery where one wins 1 pound with probability $\tfrac{2}{3}$.
Objects $x$ and $y$ are designed to be similar to each other and object $z$ is designed to be dissimilar to both $x$ and $y$.
Only the menus $\{x,z\}$, $\{y,z\}$ and $\{x,y,z\}$ are presented.
The doubleton menu $\{x,y\}$ consisting of the two similar objects is never presented.
The following code extracts the count data for choice task `Gambles` observed for experimental subject 1.

```{r data}
subject <- 1
n <- 3
N <- T_1972_counts["Gambles", subject, , ]
u <- create_universe(3, colnames(N))
print(N, na.print = '-')
Nv <- vectorize(u, N)
print(Nv)
```

The following code sets up the prior distribution of $\alpha$, a parameter of both the BRC and the BRP models, sets dimension parameters that will be passed to simulation routine, and sets a seed for pseudo-random number generation.
Given $\alpha$ and any choice menu $A \subseteq T$, the distribution of the choice probability vector for menu $A$ is symmtric Dirichlet with shape parameters all equal to $\alpha/|A|$.

```{r SMC_setup}
a <- 4      # Gamma shape parameter
b <- 0.1    # Gamma rate (inverse scale) parameter
alpha_prior <- create_alpha_prior(n, 4, 0.1)
J <- 40     # Number of independent particle groups
M <- 120    # Number of particles in each group
set.seed(123)
```

The prior distribution of the parameter $\alpha$ is $\alpha \sim \mathrm{Ga}(a,b)$ and we choose the values $a = `r a`$ and $b = `r b`$, so that the prior mean and variance of $\alpha$ are $a/b = `r a/b`$ and $a/b^2 = `r a/b^2`$.
There are $J = `r J`$ statistically independent groups, each with $M = `r M`$ particles.

## Simulation from the Bayesian random choice model

The following command performs SMC for the BRC model, returning a named list `RC_sim`.
```{r RC_sim}
RC_sim <- run_RC_sim(u, J, M, alpha_prior, Nv)
```

The most important component of `RC_sim` is `alpha`, a $M \times J$ matrix consisting of a posterior sample of the parameter $\alpha$.
By construction, the $J$ columns of `RC_sim$alpha` are mutually independent.
The function `ind_group_stats` provides a way of computing several relevant statistics for samples organized in the form of matrix with mutually independent columns.
The following code shows how to use this function to obtain the posterior mean and  standard deviation of $\alpha$, as well as the numerical standard error and the relative numerical efficiency of this mean.
```{r alpha_stats}
quant_p <- c(0.025, 0.05, 0.5, 0.95, 0.975)
stats <- ind_groups_stats(RC_sim$alpha, J, quant_p)
sprintf("Mean, standard deviation of alpha are %f, %f", stats$mean, stats$std)
sprintf("Numerical standard error of mean is %f", stats$nse)
sprintf("Relative numerical efficiency of mean is %f", stats$rne)

df <- data.frame(q = stats$q, nse = stats$q_nse)
knitr::kable(df, digits = 3, caption = "Posterior quantiles of alpha") %>%
   kableExtra::kable_styling(full_width = FALSE)
```

Another useful function, `compute_pdf_cdf_on_grid` computes probability density and cumulative density functions for samples organized as a matrix with mutually independent columns.
```{r alpha_pdf}
n_alpha_grid <- 40
RC_alpha_funcs <- compute_pdf_cdf_on_grid(RC_sim$alpha, J, n_alpha_grid)
al_pdf <- RC_alpha_funcs$pdf # cdf is also possible
plot(al_pdf$x, al_pdf$func, 'l', xlab = "alpha", ylab = "density value")
lines(al_pdf$x, al_pdf$func + al_pdf$nse, col='red')
lines(al_pdf$x, al_pdf$func - al_pdf$nse, col='red')
```

```{r binp_funcs, fig.width=6, fig.height=4}
p_grid <- seq(0, 1, by=1/80)
RC_binp_funcs <- compute_RC_binp_funcs(u, RC_sim$alpha, J, Nv, p_grid)
xy_pdf <- RC_binp_funcs[[1]]$pdf # posterior pdf function for p(x,y)
xy_pdf$upper <- xy_pdf$func + xy_pdf$nse
xy_pdf$lower <- xy_pdf$func - xy_pdf$nse
ggplot(xy_pdf, aes(x = x)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "green", alpha = 0.2) +
  geom_line(aes(y = func), color = "black") +
  labs(x = "p(x,y)", y = "density value") +
  theme_minimal()
xz_pdf <- RC_binp_funcs[[2]]$pdf # posterior pdf function for p(x,z)
xz_pdf$upper <- xz_pdf$func + xz_pdf$nse
xz_pdf$lower <- xz_pdf$func - xz_pdf$nse
ggplot(xz_pdf, aes(x = x)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "green", alpha = 0.2) +
  geom_line(aes(y = func), color = "black") +
  labs(x = "p(x,z)", y = "density value") +
  theme_minimal()
```

## SMC simulation from the Bayesian random preference model

```{r RP_sim}
lambda_values <- seq(0.01, 1.00, by=0.01)
cycle_schedule <- create_cycle_schedule(lambda_values)
RP_sim <- run_RP_sim(u, J, M, alpha_prior, Nv, lambda_values, cycle_schedule)
```


```{r ln_marl_nse, fig.width=6, fig.height=4}
# Prepare data
cycle_stats <- RP_sim$cycle_stats
cycle_stats$upper <- cycle_stats$cum_ln_marl + 2*sqrt(cycle_stats$cum_ln_marl_nse2)
cycle_stats$lower <- cycle_stats$cum_ln_marl - 2*sqrt(cycle_stats$cum_ln_marl_nse2)

ggplot(cycle_stats, aes(x = lambda)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "green", alpha = 0.2) +
  geom_line(aes(y = cum_ln_marl), color = "black") +
  labs(x = "lambda", y = "log marginal likelihood of hybrid model") +
  theme_minimal()
```

```{r ln_marl_scatter, fig.width=6, fig.height=4}
lambda_stats <- RP_sim$lambda_stats
aggr_stats <- lambda_stats$aggregates
gr_df <- as_tibble(t(lambda_stats$gr_cum_ln_marl))
gr_df$lambda <- lambda_stats$aggregates[,'lambda']

gr_df$aggr_cum_ln_marl <- aggr_stats[,'cum_ln_marl']
gr_long <- pivot_longer(
  gr_df, cols = -c(lambda, aggr_cum_ln_marl),
  names_to = "group", values_to = "cum_ln_marl"
)

ggplot() +
  geom_line(data = gr_long, aes(x = lambda, y = cum_ln_marl, group = group),
            color = "grey", size = 0.3) +
  geom_line(data = gr_long, aes(x = lambda, y = aggr_cum_ln_marl),
            color = "black", size = 1) +
  labs(x = "lambda", y = "log marginal likelihood of hybrid model") +
  theme_minimal()
```

```{r newaPr, fig.width=6, fig.height=4}
big_aPr <- as_tibble(RP_sim$big_aPr)
colnames(big_aPr) = c('x..', 'y..', 'z..')
big_aPr$lambda <- cycle_stats$lambda

aPr_long <- pivot_longer(
  big_aPr, cols = -lambda, names_to = "group", values_to = "aPr"
)

ggplot() +
  geom_line(data = aPr_long,
            aes(x = lambda, y = aPr, group = group, color = group),
            size = 0.3) +
  labs(x = "lambda", y = "Acceptance probability by block") +
  theme_minimal()
```

```{r aPr, fig.width=6, fig.height=4}
big_aPr = RP_sim$big_aPr
sm_aPr = RP_sim$sm_aPr
n_cycles <- nrow(big_aPr)
plot(seq(n_cycles), rowMeans(big_aPr), 'l',
     xlab='Cycle index l = 1,...,L',
     ylab='Acceptance probability', ylim=c(0,1))
for (i in seq(3))
  lines(seq(n_cycles), big_aPr[,i], col='grey')
lines(seq(n_cycles), rowMeans(sm_aPr), col='blue')
for (i in seq(6))
  lines(seq(n_cycles), sm_aPr[,i], col='lightgreen')
```

```{r alaPr, fig.width=6, fig.height=4}
ggplot(data = cycle_stats) +
  geom_line(aes(x = lambda, y = alpha_aPr)) +
  labs(x = "lambda", y = "Acceptance probability of alpha proposals")
```

```{r pi_graphics, echo=FALSE}
pi <- scale(RP_sim$gamma, center = F, scale = colSums(RP_sim$gamma))
pi_mean <- rowMeans(pi)
pi_cor <- cor(t(pi))
pi_thin <- pi[, seq(40, J*M, by=40)]

bp <- boxplot(t(pi_thin))
points(1:factorial(n), pi_mean)

corrplot.mixed(pi_cor, upper='ellipse', lower='number', tl.pos='lt')
```

