---
title: "SMC for random preferences"
output: rmarkdown::html_vignette
bibliography: bibliography.bib
vignette: >
  %\VignetteIndexEntry{SMC for random preferences}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
	echo = TRUE,
	message = FALSE,
	warning = FALSE,
  comment = "#>"
)
```

## Introduction

This vignette illustrates the use of tools for Sequential Monte Carlo (SMC) analysis of Bayesian random choice (BRC) and Bayesian random preference (BRP) models.
A detailed theoretical discussion of the SMC tools used here is in @McCaMarl13.
We illustrate here a small problem with a short execution time.
For an example using much more choice data, see [this Github repository][SMCRandomPref].

We first attach the required libraries.
```{r setup}
library(RanCh)
library(kableExtra)
library(corrplot)
library(tidyverse)
```

## Preparing for simulation

All of the illustrations in this vignette use data from @Tver72a, a paper that introduces both the elimination by aspects model and the "similary hypothesis", now usually known as the similary effect.
The universe of objects is the set of $n=3$ choice objects of choice task $C$ (Gambles), denoted $T = \{x,y,z\}$.
Object $x$ is the lottery where one wins 4 Israeli pounds with probability $\tfrac{1}{5}$ (and nothing otherwise);
$y$ is the lottery where one wins 3.5 pounds with probability $\tfrac{1}{4}$; and $z$ is the lottery where one wins 1 pound with probability $\tfrac{2}{3}$.
Objects $x$ and $y$ are designed to be similar to each other and object $z$ is designed to be dissimilar to both $x$ and $y$.
Only the menus $\{x,z\}$, $\{y,z\}$ and $\{x,y,z\}$ are presented.
The doubleton menu $\{x,y\}$ consisting of the two similar objects is never presented.
The following code extracts the count data for choice task `Gambles` observed for experimental subject 1.
For example, of the 20 times the menu $\{x,y\}$ was presented to this subject, the subject chose $x$ 7 times and $z$, 13 times.
The vectorized version `Nv` of the data is used in computations for the sake of effeciency.
Note that singleton choice sets are excluded.

```{r data}
subject <- 1
n <- 3
N <- T_1972_counts["Gambles", subject, , ]
u <- create_universe(3, colnames(N))
print(N, na.print = '-')
Nv <- vectorize(u, N)
print(Nv)
```

The following code sets up the prior distribution of $\alpha$, a parameter of both the BRC and the BRP models, sets dimension parameters that will be passed to simulation routines, and sets a seed for pseudo-random number generation.
Given $\alpha$ and any choice menu $A \subseteq T$, the distribution of the choice probability vector for menu $A$ is symmtric Dirichlet with shape parameters all equal to $\alpha/|A|$.

```{r SMC_setup}
a <- 4      # Gamma shape parameter
b <- 0.1    # Gamma rate (inverse scale) parameter
alpha_prior <- create_alpha_prior(n, 4, 0.1)
J <- 40     # Number of independent particle groups
M <- 120    # Number of particles in each group
set.seed(123)
```

The prior distribution of the parameter $\alpha$ is $\alpha \sim \mathrm{Ga}(a,b)$ and we choose the values $a = `r a`$ and $b = `r b`$, so that the prior mean and variance of $\alpha$ are $a/b = `r a/b`$ and $a/b^2 = `r a/b^2`$.
There are $J = `r J`$ statistically independent groups, each with $M = `r M`$ particles.

## Simulation from the Bayesian random choice model

The following command performs SMC for the BRC model, returning a named list `RC_sim`.
```{r RC_sim}
RC_sim <- run_RC_sim(u, J, M, alpha_prior, Nv)
```

The most important component of `RC_sim` is `alpha`, a $M \times J$ matrix consisting of a posterior sample of the parameter $\alpha$.
By construction, the $J$ columns of `RC_sim$alpha` are mutually independent.
The function `ind_group_stats` provides a way of computing several relevant statistics for samples organized in the form of matrix with mutually independent columns.
The following code shows how to use this function to obtain the posterior mean and  standard deviation of $\alpha$, as well as the numerical standard error and the relative numerical efficiency of this mean.
```{r alpha_stats}
quant_p <- c(0.025, 0.05, 0.5, 0.95, 0.975)
stats <- ind_groups_stats(RC_sim$alpha, J, quant_p)
sprintf("Mean, standard deviation of alpha are %f, %f", stats$mean, stats$std)
sprintf("Numerical standard error of mean is %f", stats$nse)
sprintf("Relative numerical efficiency of mean is %f", stats$rne)

df <- data.frame(q = stats$q, nse = stats$q_nse)
knitr::kable(df, digits = 3, caption = "Posterior quantiles of alpha") %>%
   kableExtra::kable_styling(full_width = FALSE)
```

Another useful function, `compute_pdf_cdf_on_grid` computes probability density and cumulative density functions for samples organized as a matrix with mutually independent columns.
The following code provides a histogram of the posterior distribution of $\alpha$ according to the BRC model.
Confidence intervals are indicated for each histogram bin, based on the independence of $J$ particle groups; the coverage probabilities are 95\%.
A kernel density is also provided.
```{r alpha_pdf, fig.width=6, fig.height=4}
n_alpha_grid <- 40
RC_alpha_funcs <- compute_pdf_cdf_on_grid(RC_sim$alpha, J, n_alpha_grid)
al_pdf <- as_tibble(RC_alpha_funcs$pdf) # cdf is also possible
al_pdf$ymin <- al_pdf$func + qnorm(0.975) * al_pdf$nse
al_pdf$ymax <- al_pdf$func + qnorm(0.025) * al_pdf$nse
bin_width <- (max(al_pdf$x) - min(al_pdf$x)) / n_alpha_grid

# Kernel density of log(alpha) transformed back to density of alpha
ln_alpha = log(RC_sim$alpha)
pdf_ln_al_kde <- density(ln_alpha, bw = "nrd0", adjust = 2)
al = exp(pdf_ln_al_kde$x)
df <- tibble(al = al, pdf = pdf_ln_al_kde$y / al)

ggplot() +
  geom_col(data = al_pdf, aes(x = x, y = func), 
           width = bin_width, fill = "skyblue", alpha = 0.6) +
  geom_errorbar(data = al_pdf, aes(x = x, ymin = ymin, ymax = ymax), 
                width = bin_width * 0.4) +
  geom_line(data = df, aes(x = al, y = pdf), color = "darkred", size = 1.2) +
  labs(x = "x", y = "Density", title = "Histogram of alpha") +
  theme_minimal()
```

The following code 
```{r binp_funcs, fig.width=6, fig.height=4}
p_grid <- seq(0, 1, by=1/80)
RC_binp_funcs <- compute_RC_binp_funcs(u, RC_sim$alpha, J, Nv, p_grid)
xy_pdf <- RC_binp_funcs[[1]]$pdf # posterior pdf function for p(x,y)
xy_pdf$upper <- xy_pdf$func + qnorm(0.995) * xy_pdf$nse
xy_pdf$lower <- xy_pdf$func + qnorm(0.005) * xy_pdf$nse
ggplot(xy_pdf, aes(x = x)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "green", alpha = 0.2) +
  geom_line(aes(y = func), color = "black") +
  labs(x = "p(x,y)", y = "density") +
  theme_minimal()
xz_pdf <- RC_binp_funcs[[2]]$pdf # posterior pdf function for p(x,z)
xz_pdf$upper <- xz_pdf$func + qnorm(0.995) * xz_pdf$nse
xz_pdf$lower <- xz_pdf$func + qnorm(0.005) * xz_pdf$nse
ggplot(xz_pdf, aes(x = x)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "green", alpha = 0.2) +
  geom_line(aes(y = func), color = "black") +
  labs(x = "p(x,z)", y = "density") +
  theme_minimal()
```

## SMC simulation from the Bayesian random preference model

The following code sets up simulation parameters that are used for simulation of the BRP model and the hybrid models, indexed by parameter $\lambda \in (0,1)$ that are intermediate between the BRC and BRP models.

```{r RP_sim}
lambda_values <- seq(0.01, 1.00, by=0.01)
cycle_schedule <- create_cycle_schedule(lambda_values)
RP_sim <- run_RP_sim(u, J, M, alpha_prior, Nv, lambda_values, cycle_schedule)
```

The following code shows the marginal likelihood of the hybrid models as a function of $\lambda$.
Error bands show confidence intervals with coverage probability 0.95.
The marginal likelihoods at $\lambda = 0$ and $\lambda = 1$ are the marginal likelihoods of the BRC and BRP models, respectively.

```{r ln_marl_nse, fig.width=6, fig.height=4}
# Prepare data
cycle_stats <- RP_sim$cycle_stats
nse = sqrt(cycle_stats$cum_ln_marl_nse2)
cycle_stats$upper <- cycle_stats$cum_ln_marl + qnorm(0.975) * nse
cycle_stats$lower <- cycle_stats$cum_ln_marl + qnorm(0.025) * nse 

ggplot(cycle_stats, aes(x = lambda)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "green", alpha = 0.2) +
  geom_line(aes(y = cum_ln_marl), color = "black") +
  labs(x = "lambda", y = "log marginal likelihood of hybrid model, by group") +
  theme_minimal()
```

The following code shows the log marginal likelihoods computed using individual partical groups.
```{r ln_marl_scatter, fig.width=6, fig.height=4}
lambda_stats <- RP_sim$lambda_stats
aggr_stats <- lambda_stats$aggregates
gr_df <- as_tibble(lambda_stats$gr_cum_ln_marl)
gr_df$lambda <- lambda_stats$aggregates[,'lambda']

gr_df$aggr_cum_ln_marl <- aggr_stats[,'cum_ln_marl']
gr_long <- pivot_longer(
  gr_df, cols = -c(lambda, aggr_cum_ln_marl),
  names_to = "group", values_to = "cum_ln_marl"
)

ggplot() +
  geom_line(data = gr_long, aes(x = lambda, y = cum_ln_marl, group = group),
            color = "grey", size = 0.2) +
  geom_line(data = gr_long, aes(x = lambda, y = aggr_cum_ln_marl),
            color = "black", size = 1) +
  labs(x = "lambda", y = "log marginal likelihood of hybrid model") +
  theme_minimal()
```

The following code shows the acceptance probability for proposals of the $\alpha$ parameter, by $\lambda$.
```{r alaPr, fig.width=6, fig.height=4}
ggplot(data = cycle_stats) +
  geom_line(aes(x = lambda, y = alpha_aPr)) +
  labs(x = "lambda", y = "Acceptance probability of alpha proposals")
```

The following code shows the acceptance probability for proposals of blocks of the $\gamma$ parameter.
The block `[x..]` of $\gamma$ weights consists of weights for the two preferences `xyz` (i.e. $x \succ y \succ z$) and `xzy`; `[y..]` consists of weights for the
preferences `yxz` and `yzx`; and `[z..]` consists of weights for the preferences `zxy` and `zyx`.
WJM: explain breaks.

```{r newaPr, fig.width=6, fig.height=4}
big_aPr <- as_tibble(RP_sim$big_aPr)
big_aPr$lambda <- cycle_stats$lambda

aPr_long <- pivot_longer(
  big_aPr, cols = -lambda, names_to = "block", values_to = "aPr"
)

ggplot() +
  geom_line(data = aPr_long,
            aes(x = lambda, y = aPr, group = block, color = block),
            size = 0.3) +
  labs(x = "lambda", y = "Acceptance probability by block") +
  theme_minimal()
```

```{r pi_graphics, echo=FALSE}
pi <- scale(RP_sim$gamma, center = F, scale = colSums(RP_sim$gamma))
pi_mean <- rowMeans(pi)
pi_cor <- cor(t(pi))
pi_thin <- t(pi[, seq(40, J*M, by=40)])

long_df <- as.tibble(t(pi)) %>%
  pivot_longer(cols = everything(), names_to = "preference", values_to = "probability")

ggplot(long_df, aes(x = probability, y = preference)) +
  geom_boxplot()

corrplot.mixed(pi_cor, upper='ellipse', lower='number', tl.pos='lt')
```

[SMCRandomPref]: https://github.com/mccauslw/SMCRandomPref
