---
title: "SMC for random preferences"
output: rmarkdown::html_vignette
bibliography: bibliography.bib
vignette: >
  %\VignetteIndexEntry{SMC for random preferences}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
	echo = TRUE,
	message = FALSE,
	warning = FALSE,
  comment = "#>"
)
```

## Introduction

This vignette illustrates the use of tools for posterior analysis of the Bayesian Dirichlet Random Choice (RC) and Bayesian random preference (RP) models using Sequential Monte Carlo (SMC) methods.
A detailed theoretical discussion of these models and the SMC tools used here is in @McCaMarl13.
We illustrate here a small problem with a short execution time.
For an example using much more choice data, see the Github repository [SMCRandomPref][SMCRandomPref].

We first attach the required libraries.

```{r setup}
library(RanCh)
library(kableExtra)
library(corrplot)
library(tidyverse)
```

## Preparing for simulation

All of the illustrations in this vignette use data from @Tver72a, a paper that introduces both the elimination by aspects model and the "similary hypothesis", now commonly known as the similary effect.
The universe of objects is the set of $n=3$ choice objects of choice task $C$ (Gambles), denoted $T = \{x,y,z\}$.
Object $x$ is the lottery where one wins 4 Israeli pounds with probability $\tfrac{1}{5}$ (and nothing otherwise);
$y$ is the lottery where one wins 3.5 pounds with probability $\tfrac{1}{4}$; and $z$ is the lottery where one wins 1 pound with probability $\tfrac{2}{3}$.
Objects $x$ and $y$ are designed to be similar to each other and object $z$ is designed to be dissimilar to both $x$ and $y$.
Only the menus $\{x,z\}$, $\{y,z\}$ and $\{x,y,z\}$ are presented.
The doubleton menu $\{x,y\}$ consisting of the two similar objects is never presented.
We first extract and display the count data for choice task `Gambles` observed for experimental subject 1.
We see, for example, that of the 20 times the menu $\{x,z\}$ was presented to this subject, the subject chose $x$ 7 times and $z$, 13 times.
The vectorized version `Nv` of the data is used in computations for the sake of effeciency.
Note that singleton choice sets are excluded from the vectorized version.

```{r data}
n <- 3
subject <- 1
N <- T_1972_counts["Gambles", subject, , ]
u <- create_universe(3, colnames(N))
print(N, na.print = '-')
Nv <- vectorize(u, N)
print(Nv)
```

We now set up the prior distribution of $\alpha$, a parameter common to both the RC and the RP models, set dimension parameters that will be passed to simulation routines, and set a seed for pseudo-random number generation.
Given $\alpha$ and any choice menu $A \subseteq T$, the distribution of the choice probability vector $P_A(\cdot)$ for menu $A$ is symmtric Dirichlet with shape parameters all equal to $\alpha/|A|$.

```{r SMC_setup}
a <- 4      # Gamma shape parameter
b <- 1.0    # Gamma rate (inverse scale) parameter
alpha_prior <- create_alpha_prior(n, a, b)
J <- 40     # Number of independent particle groups
M <- 120    # Number of particles in each group
set.seed(123)
```

The prior distribution of the parameter $\alpha$ is $\alpha \sim \mathrm{Ga}(a,b)$ and we choose the values $a = `r a`$ and $b = `r b`$, so that the prior mean and variance of $\alpha$ are $a/b = `r a/b`$ and $a/b^2 = `r a/b^2`$.
There are $J = `r J`$ statistically independent groups, each with $M = `r M`$ particles.

## Simulation from the Bayesian random choice model

Here, the `run_RC_sim` performs SMC for the RC model and returns the named list `RC_sim`.

```{r RC_sim}
RC_sim <- run_RC_sim(u, J, M, alpha_prior, Nv)
```

The most important component of `RC_sim` is `alpha`, a $M \times J$ matrix consisting of a posterior sample of the parameter $\alpha$.
By construction, the $J$ columns of `RC_sim$alpha` are mutually independent.
The function `ind_group_stats` provides a way of computing statistics for a sample organized in the form of matrix with mutually independent columns.
We use this function here to obtain the posterior mean and  standard deviation of $\alpha$, as well as the numerical standard error and the relative numerical efficiency of this mean.
Then we tabulate the results.

```{r alpha_stats}
quant_p <- c(0.025, 0.05, 0.5, 0.95, 0.975)
stats <- ind_groups_stats(RC_sim$alpha, J, quant_p)
sprintf("Mean, standard deviation of alpha are %f, %f", stats$mean, stats$std)
sprintf("Numerical standard error of mean is %f", stats$nse)
sprintf("Relative numerical efficiency of mean is %f", stats$rne)

df <- data.frame(q = stats$q, nse = stats$q_nse)
knitr::kable(df, digits = 3, caption = "Posterior quantiles of alpha") %>%
   kableExtra::kable_styling(full_width = FALSE)
```

Another function, `compute_pdf_cdf_on_grid`, computes probability density and cumulative density functions for a sample organized as a matrix with mutually independent columns.
Here we provide a histogram of the posterior distribution of $\alpha$ according to the RC model.
Confidence intervals are indicated for each histogram bin, based on the independence of $J$ particle groups; the coverage probabilities are 95\%.
A kernel density is also provided.

```{r alpha_pdf, fig.width=6, fig.height=4}
n_alpha_grid <- 40
RC_alpha_funcs <- compute_pdf_cdf_on_grid(RC_sim$alpha, J, n_alpha_grid)
al_pdf <- as_tibble(RC_alpha_funcs$pdf) # cdf is also possible
al_pdf$ymin <- al_pdf$func + qnorm(0.975) * al_pdf$nse
al_pdf$ymax <- al_pdf$func + qnorm(0.025) * al_pdf$nse
bin_width <- (max(al_pdf$x) - min(al_pdf$x)) / n_alpha_grid

# Kernel density of log(alpha) transformed back to density of alpha
ln_alpha = log(RC_sim$alpha)
pdf_ln_al_kde <- density(ln_alpha, bw = "nrd0", adjust = 2)
al = exp(pdf_ln_al_kde$x)
df <- tibble(al = al, pdf = pdf_ln_al_kde$y / al)

ggplot() +
  geom_col(data = al_pdf, aes(x = x, y = func), 
           width = bin_width, fill = "skyblue", alpha = 0.6) +
  geom_errorbar(data = al_pdf, aes(x = x, ymin = ymin, ymax = ymax), 
                width = bin_width * 0.4) +
  geom_line(data = df, aes(x = al, y = pdf), color = "darkred", size = 1.2) +
  labs(x = "x", y = "Density", title = "Histogram of alpha") +
  theme_minimal()
```

The function `compute_RC_binp_funcs` provides the pdf and cdf, with numerical standard errors, of all binary choice probabilities.
It integrates out $\alpha$ from the joint posterior distribution of $\alpha$ and the binary choice probabilities.
Because this is a univariate integration, and there is modest posterior correlation between $\alpha$ and the binary choice probabilities, the numerical standard error is very low.
The following code gives the posterior density of the binary choice probability $p(x,z)$, the probability of choosing $x$ from the menu $\{x,z\}$.
Error bands in green show confidence intervals with coverage probability $0.95$, based on independent groups.
While a confidence region, based on a coverage probability of 95\% and the independence of $J$ particle groupes, is plotted, the intervals are too narrow to be visible.

```{r binp_funcs, fig.width=6, fig.height=4}
p_grid <- seq(0, 1, by=1/80)
RC_binp_funcs <- compute_RC_binp_funcs(u, RC_sim$alpha, J, Nv, p_grid)
xz_pdf <- RC_binp_funcs[[2]]$pdf # posterior pdf function for p(x,z)
xz_pdf$upper <- xz_pdf$func + qnorm(0.975) * xz_pdf$nse
xz_pdf$lower <- xz_pdf$func + qnorm(0.025) * xz_pdf$nse
ggplot(xz_pdf, aes(x = x)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "green", alpha = 0.2) +
  geom_line(aes(y = func), color = "black", linewidth = 0.2) +
  labs(x = "p(x,z)", y = "density") +
  theme_minimal()
```

## SMC simulation from the Bayesian random preference model

Here we set up simulation parameters that are used for simulation of the RP model and the hybrid models, indexed by parameter $\lambda \in (0,1)$, that are intermediate between the RC and RP models.
The user supplies a grid of values of lambda, `lambda_values`, at which various quantities are evaluated.
The function `create_cycle_schedule()` returns a grid, coarser than `lambda_values` and irregularly spaced, of $\lambda$ values defining the SMC cycles, along with SMC parameters that vary by cycle.

```{r RP_sim}
lambda_values <- seq(0.01, 1.00, by=0.01)
cycle_schedule <- create_cycle_schedule(lambda_values)
RP_sim <- run_RP_sim(u, J, M, alpha_prior, Nv, lambda_values, cycle_schedule)
```

Next we plot the marginal likelihood of the hybrid models as a function of $\lambda$.
Here, the marginal likelihood is computed on the grid of values of $\lambda$ returned by create_cycle_schedule, `cycle_stats$lambda`.
Error bands show confidence intervals with coverage probability 0.95, based on independent groups.
The marginal likelihoods at $\lambda = 0$ and $\lambda = 1$ are the marginal likelihoods of the RC and RP models, respectively.

```{r ln_marl_nse, fig.width=6, fig.height=4}
# Prepare data
cycle_stats <- RP_sim$cycle_stats
nse = sqrt(cycle_stats$cum_ln_marl_nse2)
cycle_stats$upper <- cycle_stats$cum_ln_marl + qnorm(0.975) * nse
cycle_stats$lower <- cycle_stats$cum_ln_marl + qnorm(0.025) * nse 

ggplot(cycle_stats, aes(x = lambda)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "green", alpha = 0.2) +
  geom_line(aes(y = cum_ln_marl), color = "black", linewidth = 0.2) +
  labs(x = "lambda", y = "log marginal likelihood of hybrid model, by group") +
  theme_minimal()
```

The following code shows the log marginal likelihoods computed using individual partical groups.

```{r ln_marl_scatter, fig.width=6, fig.height=4}
lambda_stats <- RP_sim$lambda_stats
aggr_stats <- lambda_stats$aggregates
gr_df <- as_tibble(lambda_stats$gr_cum_ln_marl)
gr_df$lambda <- lambda_stats$aggregates[,'lambda']

gr_df$aggr_cum_ln_marl <- aggr_stats[,'cum_ln_marl']
gr_long <- pivot_longer(
  gr_df, cols = -c(lambda, aggr_cum_ln_marl),
  names_to = "group", values_to = "cum_ln_marl"
)

ggplot() +
  geom_line(data = gr_long, aes(x = lambda, y = cum_ln_marl, group = group),
            color = "grey", size = 0.2) +
  geom_line(data = gr_long, aes(x = lambda, y = aggr_cum_ln_marl),
            color = "black", size = 1) +
  labs(x = "lambda", y = "log marginal likelihood of hybrid model") +
  theme_minimal()
```

Here we illustrate the posterior distribution of $\pi$ for the RP model.
The vector $\pi$ is a vector giving the probabilities of the six preference orders on $T = \{x,y,z\}$.

```{r pi_graphics, echo=FALSE}
pi <- scale(RP_sim$gamma, center = F, scale = colSums(RP_sim$gamma))
pi_cor <- cor(t(pi))
order_names <- rownames(pi)

long_df <- as.tibble(t(pi)) %>%
  pivot_longer(cols = everything(),
               names_to = "preference", values_to = "probability")

ggplot(long_df, aes(x = probability, y = fct_rev(preference))) +
  geom_boxplot() +
  theme_minimal()

corrplot.mixed(pi_cor, upper='ellipse', lower='number', tl.pos='d')
```

## Diagnostics for the SMC simulations for the RP model

Here we plot the acceptance probability for proposals of the $\alpha$ parameter, at the M (mutation) phase of each SMC cycle.
We plot it against the value of $\lambda$ prevailing at the cycle.
```{r alaPr, fig.width=6, fig.height=4}
ggplot(data = cycle_stats) +
  geom_point(aes(x = lambda, y = alpha_aPr)) +
  labs(x = "lambda", y = "Acceptance probability of alpha proposals")
```
Notice the uneven grid of points defining the SMC cycles.
We need more simulation for smaller and larger values of $\lambda$ to deal with particle depletion.

Next, we show the acceptance probability for proposals of blocks of the $\gamma$ parameter, also at the M phase of each SMC cycle.
The block `[x..]` of $\gamma$ weights consists of weights for the two preferences `xyz` (i.e. $x \succ y \succ z$) and `xzy`; `[y..]` consists of weights for the
preferences `yxz` and `yzx`; and `[z..]` consists of weights for the preferences `zxy` and `zyx`.
```{r newaPr, fig.width=6, fig.height=4}
big_aPr <- as_tibble(RP_sim$big_aPr)
big_aPr$lambda <- cycle_stats$lambda

aPr_long <- pivot_longer(
  big_aPr, cols = -lambda, names_to = "block", values_to = "aPr"
)

ggplot() +
  geom_point(data = aPr_long,
             aes(x = lambda, y = aPr, group = block, color = block),
             size = 0.3) +
  labs(x = "lambda", y = "Acceptance probability by block") +
  theme_minimal()
```


[SMCRandomPref]: https://github.com/mccauslw/SMCRandomPref
